---
title: "ETC5242 Assignment 1 - Group 31"
author: "Ma Sun Yan Joanna 33775877" 
date: "2024-10-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

```{r library}

# Library
library(tidyverse)
library(kableExtra)

```


This report conducts statistic analysis of bank customers, based on a database including data of 200 customers' annual gross income and education level.

# Task 1

First part would be some distribution plots and descriptive statistics of the data, and fitted distribution model.  

```{r read_data}

# Read in data
bank <- read_csv(here::here("data/banksurvey.csv"))

```

```{r desc_plot}
# Descriptive bar plot
bank %>% 
  group_by(education) %>% 
  summarise(perc = n()/nrow(bank) * 100) %>% 
  as.data.frame() %>% 
  ggplot(aes(x = fct_reorder(education, -perc), 
             y = perc)) + 
  geom_col() + 
  theme_bw() + 
  labs(x = "Education Level", 
       y = "Percentage", 
       title = "Education level of bank customers") + 
  geom_label(aes(label = paste0(perc, "%"), 
                 vjust = 0.5))

```

The bar plot shows that most of the surveyed bank customers achieved a highest education level of bachelor's degree (66%), followed by master's and high school with similar percentage, and 2% hold a doctorate degree.  

```{r desc_stat}

# Descriptive statistics
bank %>% 
  summarise(Min = min(income), 
            Q1 = round(quantile(income, 0.25), 1), 
            Median = round(median(income), 1), 
            Mean = round(mean(income), 1), 
            Q3 = round(quantile(income, 0.75), 1), 
            Max = max(income),
            sd = sd(income)) %>% 
  kable(caption = "Annual gross income of bank customers (in thousands)")

```

The above table shows the descriptive statistics of the annual gross income in the previous financial year


Explanation what analysis
results plots numerical, conclusion
Details about analysis choice and assumption for statistican
summary of method and findings and key assumptions for manager

# Task 2

In this section, it will focus on estimating the 80th percentile of annual income of customers. In addition, 95% confidence intervals based on each of the estimators will be calculated.

```{r Task2-1}
# Estimate the 80th percentile using estimators based on fitted model
q80_normal <- qnorm(0.80, mean = fit_norm$estimate[1], sd = fit_norm$estimate[2])
q80_exponential <- qexp(0.80, rate=fit_exp$estimate)
q80_gamma <- qgamma(0.80, shape=fit_gamma$estimate[1], rate=fit_gamma$estimate[2])

# Create a tibble based on these results
results <- tibble(
  Distribution = c("Normal", "Exponential", "Gamma"),
  `80th Percentile Estimate` = c(q80_normal, q80_exponential, q80_gamma))


kable(results, caption = "80th Percentile Estimates based on Different Models")
```

The parameter of interest is the 80th percentile of the annual income distribution. It indicates the income level at which 80% of the customers earn less than this amount.

Although we can estimate the 80th percentile of annual income based on a particular model, it is also possible to use a statistic called 'sample quantile', which is an estimator that does not assume any specific model.

```{r Task2-2}
# Using sample quantile to estimate
q80_sample <- quantile(data$income, probs = 0.80)

new_results <- tibble(
  Distribution = c("Normal", "Exponential", "Gamma","Sample"),
  `80th Percentile Estimate` = c(q80_normal, q80_exponential, q80_gamma,q80_sample))


kable(new_results, caption = "80th Percentile Estimates based on Different Models")
```

```{r Task2-3}
# Calculate the 95% CI for the estimates

# Normal distribution 95% CI
se_normal <- fit_norm$estimate[2] / sqrt(length(data$income))  
ci_normal <- q80_normal + c(-1.96, 1.96) * se_normal

# Exponential distribution 95% CI
se_exp <- sqrt(1 / (length(data$income) * fit_exp$estimate^2))
ci_exp <- q80_exponential + c(-1.96, 1.96) * se_exp

# Gamma distribution 95% CI
se_gamma <- sqrt(fit_gamma$estimate[1] / (fit_gamma$estimate[2]^2 * length(data$income)))
ci_gamma <- q80_gamma + c(-1.96, 1.96) * se_gamma

ci_results <- tibble(
  Distribution = c("Normal", "Exponential", "Gamma"),
  `80th Percentile Estimate` = c(q80_normal, q80_exponential, q80_gamma),
  `Lower Bound (95% CI)` = c(ci_normal[1], ci_exp[1], ci_gamma[1]),
  `Upper Bound (95% CI)` = c(ci_normal[2], ci_exp[2], ci_gamma[2])
)

kable(ci_results, caption = "80th Percentile Estimates and 95% Confidence Intervals for Different Distributions")
```

# Task 3



# Task 4

# Task 5