---
title: "ETC5242 Assignment 1 - Group 31"
author: "Ma Sun Yan Joanna 33775877"
date: "2024-10-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

```{r library}

# Library
library(tidyverse)
library(kableExtra)

```


This report conducts statistic analysis of bank customers, based on a database including data of 200 customers' annual gross income and education level.

# Task 1

First part would be some distribution plots and descriptive statistics of the data, and fitted distribution model.  

```{r read_data}

# Read in data
bank <- read_csv(here::here("data/banksurvey.csv"))

```

```{r desc_plot}

# Descriptive bar plot
bank %>% 
  group_by(education) %>% 
  summarise(perc = n()/nrow(bank) * 100) %>% 
  as.data.frame() %>% 
  ggplot(aes(x = fct_reorder(education, -perc), 
             y = perc)) + 
  geom_col() + 
  theme_bw() + 
  labs(x = "Education Level", 
       y = "Percentage", 
       title = "Education level of bank customers") + 
  geom_label(aes(label = paste0(perc, "%"), 
                 vjust = 0.5))

```

The bar plot shows that most of the surveyed bank customers achieved a highest education level of bachelor's degree (66%), followed by master's and high school with similar percentage, and 2% hold a doctorate degree.  

```{r desc_stat}

# Descriptive statistics
bank %>% 
  summarise(Min = min(income), 
            Q1 = round(quantile(income, 0.25), 1), 
            Median = round(median(income), 1), 
            Mean = round(mean(income), 1), 
            Q3 = round(quantile(income, 0.75), 1), 
            Max = max(income),
            sd = sd(income)) %>% 
  kable(caption = "Annual gross income of bank customers (in thousands)")

```

The above table shows the descriptive statistics of the annual gross income in the previous financial year


Explanation what analysis
results plots numerical, conclusion
Details about analysis choice and assumption for statistican
summary of method and findings and key assumptions for manager

# Task 2

# Task 3

# Task 4

# Task 5